\pkg{ggenealogy}: An \proglang{R} Package for Visualizing Genealogical Data

### 2016-03-10 JSS/jss-style/article.tex
Generalize intro outside of biological context

%% an abstract and keywords
\Abstract{
  This paper introduces \pkg{ggenealogy} (\citealt{ggen}), a developing \proglang{R} package that provides tools for searching through genealogical data, generating basic statistics on their graphical structures using parent and child connections, and displaying the results. It is possible to draw the genealogy in relation to variables related to the nodes, and to determine and display the shortest path distances between genetic lines. Production of pairwise distance matrices and genealogical diagrams constrained on generation are also available in the visualization toolkit. The software is being tested on datasets with milestone cultivars of soybean (\citealt{soybean}) and barley varieties [2], as well as on data from the \citealt{mgp}, a database of the academic genealogy of mathematicians. It is currently available on the Comprehensive R Archive Network.}

@@

Genealogy is the study of the parent-child relationships. By tracing through lineages of groups of individuals, the history of features that have been modified over time can be studied. Comparative geneticists, computational biologists, and bioinformaticians commonly use these tools to better understand the historical changes that caused novel and desirable traits to arise in lineages. For example, in crops, desirable modifications could include an increase in protein yield or an increase in disease resistance. However, there are also times when lineages of detrimental traits are analyzed, such as to determine the origin of hazardous traits in rapidly-evolving viruses.

Furthermore, genealogical relationships can be applied outside of a strict biological sense, an example being the \citealt{mgp}, which maintains the family lineage of academic mathematicians, with documentation of doctoral advisors, doctoral students, and graduation years of all mathematicians in academia. Such family lineages allow us to understand the position of one member in the larger historical picture, and to accurately preserve past relationships for the knowledge of future generations.

In all these examples, the data structures containing the genealogical relationships can be represented visually. Access to various types of visual plots and diagrams of the lineage can allow scientists and others to more efficiently and accurately explore an otherwise complicated data structure. We introduce here a developing visualization toolkit that is intended to assist users in their exploration and analysis of genealogical relationships. In this paper, we will demonstrate some of the main features of this software package, and summarize any novelty that such features may provide users, using an example lineage dataset of soybean cultivars (\citealt{soybean}).

>>>>

%% an abstract and keywords
\Abstract{
  This paper introduces \pkg{ggenealogy} (\citealt{ggen}), a developing \proglang{R} software package that provides tools for searching through genealogical data, generating basic statistics on their graphical structures using parent and child connections, and displaying the results. It is possible to draw the genealogy in relation to variables related to the nodes, and to determine and display the shortest path distances between the nodes. Production of pairwise distance matrices and genealogical diagrams constrained on generation are also available in the visualization toolkit. The tools are being tested on a dataset with milestone cultivars of soybean varieties (\citealt{soybean}) as well as on a web-based database of the academic genealogy of mathematicians (\citealt{mgp}). The software package is currently available on the Comprehensive R Archive Network.}

@@

Genealogy is the study of parent-child relationships. By tracing through parent-child lineages, genealogists can study the histories of features that have been modified over time. Comparative geneticists, computational biologists, and bioinformaticians commonly use genealogical tools to better understand the histories of novel traits arising across biological lineages. For example, desirable modifications in crops could include an increase in protein yield or an increase in disease resistance, and genealogical structures could be used to assess how these desirable traits developed. At the same time, genealogical lineages can also be used to assess detrimental features, such as to determine the origin of hazardous traits in rapidly-evolving viruses.

Genealogical structures can also serve as informative tools outside of a strict biological sense. For instance, we can trace mentoring relationships between students and disseration supervisors with the use of academic genealogies. This can allow us to understand the position of one member in the larger historical picture, and to accurately preserve past relationships for the knowledge of future generations. Similarly, linguistic genealogies can be used to decipher the historical changes of vocabulary and grammatical features across related languages. In short, there is a diverse array of disciplines that can elicit useful information about features of interest by using genealogical data.

In all these examples, the genealogical relationships can be represented visually. Access to various types of plotting tools can allow scientists and others to more efficiently and accurately explore features of interest across the genealogy. We introduce here a developing visualization toolkit that is intended to assist users in their exploration and analysis of genealogical structures. In this paper, we will demonstrate the main tools of this software package using two example genealogical datasets, one of soybean cultivars (\citealt{soybean}) and the other of academic mathematicians (\citealt{mgp}).
